Size convergence of the \boldmath{$\exb$} staircase pattern in flux tube simulations of ion temperature gradient driven turbulence

### 2023-05-21 bachelorthesis/41-resolution.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without fearing numerical dissipation. Numerical dissipatation can therefore result to no formation of zonal flow stuctures, which cause an permanent turbulent state of the simulation. The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation. \\
Because of that, the criteria for the best resolution should be:
\begin{enumerate}
	\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
	\item[\textbf{(2)}] Stability for \textbf{long} time periods 
\end{enumerate}
and the following procedure will be applied for verification:
\begin{enumerate}
	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
	\item Reduce to knowm minimum number of grid points to verify result in general.
\end{enumerate}

@@

Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab. \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref. \citenum{Peeters2016} in Section IV. Note that, because of selected circular cencentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref. \citenum{Peeters2016} for $s$-$\alpha$ geometry. In Fig. \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref. \citenum{Peeters2016} with subdued turbulence after $\sim 3000 R/\vth$.

\includegraphicsHere{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
}{fig:eflux-ns16-nvpar64-nmu9}{}

As next step an approach to the finite heat flux threshold were made to verify the selection of the gradient length $\rlt$. As in Ref. \citenum{Peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG. 4 of Ref. \citenum{Peeters2016}]. Therefore following parameters were used:
\begin{gather*}
	\rlt \in [6.0, 6.3]~.
\end{gather*}
As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.

\includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
}{fig:eflux-gradient-length}{}

@@

In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to:
\begin{gather*}
	\Nvpar \in [16, 32, 48, 64]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visable that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period. But resolution $\Nvpar = 32$ is as well not acceptable according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000 R/\vth$.\\
So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be:
\begin{gather*}
	\boxed{\Nvpar = 48}
\end{gather*}
With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500 R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced parallel velocity grid points $\Nvpar$
}{fig:eflux-ns16-nvpar16-32-48-64-nmu9}{}

@@

As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with:
\begin{gather*}
	\Nmu \in [6, 9] ~.
\end{gather*}
As in Fig. \ref{fig:eflux-ns16-nvpar64-nmu6-9} shown, the reduction of grid points for the magnetic moment does not significantly impact the suppression of turbulence. The turbulence enters the stationary state in both cases after $\sim 3000 R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar64-nmu6-9}{}

To conclude a curial result the number of grids point for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}. In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution:
\begin{gather*}
	\boxed{\Nmu = 9}
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ and $\Nvpar = 48$ for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar48-nmu6-9}{}

@@

In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters:
\begin{gather*}
	\Ns \in [12, 16]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns12-16-nvpar64-nmu9} is clearly visible that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the the stationary state of the turbulence gets instabil after $\sim 2500 R/\vth$. This concludes to the following resolution:
\begin{gather*}
	\boxed{\Ns = 16}
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for reduced magnetic field grid points $\Ns$
}{fig:eflux-ns12-16-nvpar64-nmu9}{}

>>>>

At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without fearing numerical dissipation. Numerical dissipatation can therefore result to no formation of zonal flow stuctures, which cause an permanent turbulent state of the simulation. The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation. \\
Because of that, the criteria for the best resolution should be:
\begin{enumerate}
	\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
	\item[\textbf{(2)}] Stability for \textbf{long} time periods 
\end{enumerate}
and the following procedure will be applied for verification:
\begin{enumerate}
	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
	\item Reduce to known minimum number of grid points to verify result in general.
\end{enumerate}

@@

Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab. \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref. \citenum{Peeters2016} in Section IV. Note that, because of selected circular cencentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref. \citenum{Peeters2016} for $s$-$\alpha$ geometry. In Fig. \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref. \citenum{Peeters2016} with subdued turbulence after $\sim 3000\,R/\vth$.

\includegraphicsHere{S6_rlt6.0/boxsize1x1/Ns16/Nvpar64/Nmu9/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux.pdf}{
	Time trace of the heat conduction coefficient $\chi$ for $\rlt = 6.0$ for benchmark
}{fig:eflux-ns16-nvpar64-nmu9}{}

As next step an approach to the finite heat flux threshold were made to verify the selection of the gradient length $\rlt$. As in Ref. \citenum{Peeters2016} in Section V conclude is the finite heat flux threshold approximately located at a grendent length of $\rlt = 6.3$ for circular geometry [FIG. 4 of Ref. \citenum{Peeters2016}]. Therefore following parameters were used
\begin{gather*}
	\rlt \in [6.0, 6.3]~.
\end{gather*}
As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.

\includegraphicsHere{Comparison/Gradient-Length/S6_rlt6.0-6.3_boxsize1x1_Ns16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces for $\rlt = 6.0$ and $\rlt = 6.3$ for benchmark
}{fig:eflux-gradient-length}{}

@@

In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to
\begin{gather*}
	\Nvpar \in [16, 32, 48, 64]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visible that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period. But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution will be
\begin{gather*}
	\boxed{\Nvpar = 48}~.
\end{gather*}
With this number of grid points the time till turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nvpar/S6_rlt6.0_boxsize1x1_Ns16_Nvpar16-32-48-64_Nmu9_eflux_comparison.pdf}{
	Time traces for reduced parallel velocity grid points $\Nvpar$
}{fig:eflux-ns16-nvpar16-32-48-64-nmu9}{}

@@

As next step the number of grid points for the magnetic moment $\Nmu$ were reduced with
\begin{gather*}
	\Nmu \in [6, 9] ~.
\end{gather*}
As in Fig. \ref{fig:eflux-ns16-nvpar64-nmu6-9} shown, the reduction of grid points for the magnetic moment does not significantly impact the suppression of turbulence. The turbulence enters the stationary state in both cases after $\sim 3000\,R/\vth$.

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar64_Nmu6-9_eflux_comparison.pdf}{
	Time traces for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar64-nmu6-9}{}

To conclude a curial result the number of grids point for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}. In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with the both criteria in mind, to the following resolution
\begin{gather*}
	\boxed{\Nmu = 9}~.
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Nmu/S6_rlt6.0_boxsize1x1_Ns16_Nvpar48_Nmu6-9_eflux_comparison.pdf}{
	Time traces for reduced magnetic moment grid points $\Nmu$
}{fig:eflux-ns16-nvpar48-nmu6-9}{}

@@

In the final step the number of grid points for the magnetic field $\Ns$ get reduced with the following parameters
\begin{gather*}
	\Ns \in [12, 16]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns12-16-nvpar64-nmu9} can be seen that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the stationary state of the turbulence gets instabil after $\sim 2500\,R/\vth$. This concludes to the following resolution
\begin{gather*}
	\boxed{\Ns = 16}~.
\end{gather*}

\includegraphicsHere{Comparison/Resolution/Ns/S6_rlt6.0_boxsize1x1_Ns12-16_Nvpar64_Nmu9_eflux_comparison.pdf}{
	Time traces for reduced magnetic field grid points $\Ns$
}{fig:eflux-ns12-16-nvpar64-nmu9}{}

### 2023-05-21 bachelorthesis/50-closure.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

>>>>

In this thesis the minimal resolution for simulations with GKW in the Cyclone Base parameter were observed in which the number of grid points for the parallel velocity $\Nvpar$ could be reduced from 64 to 48 which halfs the time until suppression. \bigskip

Additional the active development of a restart script in \texttt{python} leads to further convenience during the task of performing a simulation on the \texttt{btrzx1} cluster. \bigskip

### 2023-05-21 briefcommunication/briefcommunication.tex
#39: Correction by mayself, #37: Whitspace between formula and sign

\title[Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
{Size convergence of the $\exb$ staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}

@@

In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76,~89)\,\rhoth$ in the radial and binormal direction. Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate. Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.
The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size. \\
The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Rath2016, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2},
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0).
\end{equation}
The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\
Consistent with Ref.~\onlinecite{Peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass). Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref. \onlinecite{Peeters2016,Peeters2009}.
\newpage
In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
It is defined by
\begin{equation}
	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord),
	\label{eq:shearingrate_fourier}
\end{equation}
where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
% define the zonal flow modes' amplitude in terms of the shearing rate
Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
% introduces the term basic mode of the pattern for later usage
In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\

>>>>

\title[Size convergence of the \boldmath{$\exb$} staircase pattern in flux tube simulations of ion temperature gradient driven turbulence]
{Size convergence of the \boldmath{$\exb$} staircase pattern in flux tube simulations of ion temperature gradient driven turbulence}

@@

In the following the box size is increased relative to the standard box size $(L_\xcoord,~L_\ycoord) = (76,~89)\,\rhoth$ in the radial and binormal direction. Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate. Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}.
The increased box sizes are indicated by the real parameter $\NR$ for radial and $\NB$ for the binormal direction with the nomenclature $\NR\times \NB$ throughout this work.
Note that, the number of modes in the respective direction, i.e., $N_x$ and $N_m$, respectively, is always adapted accordingly to retain a spatial resolution compliant to the standard resolution [Tab.~\ref{tab:resolution}] and standard box size. \\
The $\exb$ staircase pattern is manifest as radial structure formation in the $\exb$ shearing rate defined by\cite{Rath2016, Pueschel2008, Rath2016, Peeters2016}
\begin{equation}
	\wexb = \frac{1}{2} \frac{\partial^2 \langle \phi \rangle}{\partial \xcoord^2}~,
	\label{eq:shearingrate}
\end{equation}
where $\langle \phi \rangle$ is the zonal electrostatic potential normalized by $\rho_\ast T/e$ ($\rho_\ast = \rhoth/R$ is the thermal Larmor radius normalized with the major radius $R$, $T$ is the temperature, $e$ is the elementary charge).
The zonal potential is calculated from the electrostatic potential $\phi$ on the two-dimensional $\xcoord$-$\ycoord$-plane at the low field side according to\cite{Rath2021}
\begin{equation}
\langle \phi \rangle = \frac{1}{L_\ycoord} \int_0^{L_\ycoord} \mathrm{d}\ycoord ~ \phi(\xcoord,\ycoord,s=0)~.
\end{equation}
The E$\times$B shearing rate $\wexb$ is the radial derivative of the advecting zonal flow velocity \cite{Hahm1995, Waltz1998} and quantifies the zonal flow induced shearing of turbulent structures \cite{Biglari1990, Hahm1995, Burnell1997}. \\
Consistent with Ref.~\onlinecite{Peeters2016} the turbulence level is quantified by the turbulent heat conduction coefficient $\chi$, which is normalized by $\rhoth^2 \vth/R$ ($\vth = \sqrt{2 T/m}$ is the thermal velocity and $m$ is the mass). Furthermore, quantities $\rhoth$, $R$, $T$, $\vth$ and $m$ are referenced quantities from Ref. \onlinecite{Peeters2016,Peeters2009}.
\newpage
In order to diagnose the temporal evolution of the staircase pattern and to obtain an estimate of its amplitude the radial Fourier transform of the $\exb$ shearing rate is considered. 
It is defined by
\begin{equation}
	\wexb = \sum_{\kzf} \hatwexb(\kzf,t) \, \exp(\mathrm{i} \kzf \xcoord)~,
	\label{eq:shearingrate_fourier}
\end{equation}
where $\hatwexb$ is the complex Fourier coefficient and \linebreak $\kzf = 2\pi \nzf/L_\xcoord$
% explicitly introduce the terms zonal flow wave vector and zonal flow mode number for later usage
defines the zonal flow wave vector with the zonal flow mode number $\nzf$ ranging in $-(N_\xcoord -1)/2 \leq \nzf \leq (N_\xcoord -1)/2 $.
% define the zonal flow modes' amplitude in terms of the shearing rate
Based on the definitions above, the shear carried by the zonal flow mode with wave vector $\kzf$ is defined by $\hatwexbamp = 2 |\hatwexb(\kzf,t)|$. 
% introduces the term basic mode of the pattern for later usage
In general, the zonal flow mode that dominates the $\exb$ staircase pattern, also referred to as the \textit{basic mode} of the pattern in this work, exhibits the maximum amplitude in the spectrum $\hatwexbamp$.\\
